---
title: "Exercise 2: Tidy Data"
author: "Eric Kelchlin"
date: "March 2019"
output: word_document
   
---
<style type="text/css">

body, td {
   font-size: 16pt;
}
code.r{
  font-size: 16pt;
}
pre {
  font-size: 16pt
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")
```

# Load the required packages

You can use requre() or library() to load a package.

```{r, required packages,message=FALSE, warning=FALSE}

require(knitr)      # to knit our documents
require(tidyverse)  # to wrangle data
require(readxl)     # to import excel sheets
require(kableExtra) # for nice tables
require(here)

```

## Task 1.1 - Import and view the data

The easiest way to read-in Excel data is to use the `read_excel()` function from the readxl package which is a happy member of the tidyverse family of packages.

```{r, Importing Data}

Ibis <- read_excel(here("Ibis_Data.xlsx"))

```


What if we had a spreadsheet with a couple lines that we wanted to skip or perhaps a specific range of cells that we only wanted to read-in?  Use the help on the `read_excel()` function to see if it's possible. Do this by simply clicking *F1* while your cursur is on the function name.

# View the Data

```{r, viewing the data}

glimpse(Ibis) 

names(Ibis)

head(Ibis, n = 5)

View(Ibis) # note the capital V

```


```{r, To knit the table in a nice format, eval = FALSE}

# Remove the eval = FALSE and run code chunk

kable(Ibis, caption = "Table 1: Ibis Data in a Wide Format") %>%
  kable_styling(full_width = FALSE, font_size = 16)
   
```

# Task 1.2 - Convert the table from a wide to a long format ##

Use the pipe **%>%** symbol **(Ctrl + Shft + M)** to simplify coding. 

### gather() function template

new object <- data %>% 
  gather(key = , value = , columns to gather, other arguments)
  
```{r, Task 1.2 - wide to long format}

# Remove the eval = FALSE and run code chunk

Ibis2 <- Ibis %>% 
  gather(key = MyYear, value = MyCount, `1972`:`1975`, convert = TRUE)

# What does the convert = TRUE argument do?

# Knit the table in a nice format. Notice how you can nest functions with the pipe.

Ibis2 %>% 
  head(n=10) %>% 
  kable(caption="Table 2: Ibis Data in a Long Format") %>%
  kable_styling(full_width = T, bootstrap_options = c("hover", "condensed"))

```

## Task 1.3 - There are two other ways to identify the the fields to gather in this example. Can you figure this out?  Tip: one way is to identify what not to gather and the other identifies the columns by location.

```{r, Task 1.3} 

Ibis3 <- Ibis %>% 
  gather(key = MyYear, value = MyCount, -(1:4), convert = TRUE) # or use names of fields

Ibis4 <- Ibis %>% 
  gather(key = MyYear, value = MyCount, 5:8, convert = TRUE)

```

## Task 1.3 - The `spread()` function restructures the long format to a wide format.  Use this function to return our gathered dataset back into the orignal format.  This one is easier than the gather function.

###  spread() template

newobject <- data %>% 
  spread(key = column to pivot, value = column with the values)

```{r, Task 3}  

IbisWide <- Ibis2 %>% spread(MyYear, MyCount)

```

## Task 1.5 - Export the Data. Run the line below to easily export the data as a .csv file. Find the `write_?` function from the readr package to export the data as a comma delimited (" ") .txt file.  

```{r, write the tables out as a csv, message = FALSE, eval=FALSE}

write_csv(Ibis2, here("IbisClean.csv"))

write_delim(Ibis2, here("IbisClean.txt"))

```

## Task 1.6 - Independent Exercise

1.  Use the `read_csv()` function to import the *Sage Grouse Data.csv* dataset.  Assign the data into a new object called *Sage*

2.  Examine the data.  How many Observations and variables do we have? What are the variable names?  What data type are the 4 date fields and what should they be?

3. We have multiple observational units combined in one table. How many do you think we have?

4.  Some variables are in wide format (i.e., observations), can you pick them out? 

```{r, Task 1.6 }
Sage <- read_csv("Sage Grouse Data.csv")
head(Sage, n = 5)
glimpse(Sage)
```

# Task 1.7 - Class Exercise

Question: What is the maximum count of male and female sage grouse per lek?

To answer this question, we'll need to use the `select()`, `group_by()` and `summarise()`functions from the **dplyr** package.  This is a great introduction of what we will be learning more of tomorrow.

## Steps:

1. Select only the variables you need using the `select()` function.  
2. Gather the *FieldResults* fields into a long format using the `gather()` function.  Use the `na.rm = TRUE` argument to remove the cells with missing values.
3. Seperate the field with the data using the `seperate()` function
4. Use the `group_by()` and `summarise()`functions to calculate the maximum number of sage grouse per lek.  Note:  `summarise()` is finiky to missing values, that's why we need to filter them out first.


```{r, Task 1.7}
Sage1 <- Sage %>% 
  select(LOCATION_C, FieldResults1:FeildResults4) %>% 
  gather(key = Results, value = SexCount, -LOCATION_C, na.rm = TRUE) %>% 
  separate(SexCount, c("No_Males", "MSex", "No_Females", "FSex"), sep = " ") %>% 
  select(LOCATION_C, No_Males, No_Females)

Sage2 <- Sage1 %>% 
  group_by(LOCATION_C) %>% 
  summarise(max_count_male = max(No_Males), max_count_female = max(No_Females))

```









