---
title: "Exercise 2: Tidy Data"
author: "Eric Kelchlin"
date: "May 2018"
output: html_document
   
---
<style type="text/css">

body, td {
   font-size: 16pt;
}
code.r{
  font-size: 16pt;
}
pre {
  font-size: 16pt
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")
```

# Load the required packages

You can use require() or library() to load a package.

```{r, required packages,message=FALSE, warning=FALSE}

require(knitr)      # to knit our documents
require(tidyverse)  # to wrangle data
require(readxl)     # to import excel sheets
require(kableExtra) # for nice tables

```

# Set your working directory

Edit the path to fit your configuration.  This is an easy way to ensure that any outputs are saved to a known location.

```{r, set directory, message = FALSE}

setwd("C:/CSP4200/Exercise3")

```

# Read-in Excel data

The easiest way to read-in Excel data is to use the `read_excel()` function from the readxl package which is a happy member of the tidyverse family of packages.

```{r, Importing Data}

Ibis <- read_excel("Ibis_Data.xlsx", sheet = "Ibis_Data")

```

## Task 1:

What if we had a spreadsheet with a couple lines that we wanted to skip or perhaps a specific range of cells that we only wanted to read-in?  Use the help on the `read_excel()` function to see if it's possible. Do this by simply clicking *F1* while your cursur is on the function name.

# View the Data

```{r, viewing the data}

glimpse(Ibis) 

names(Ibis)

head(Ibis, n = 5)

View(Ibis) # note the capital V

```

```{r, To knit the table in a nice format}

kable(Ibis, caption = "Table 1: Ibis Data in a Wide Format") %>%
  kable_styling(full_width = FALSE, font_size = 16)
   
```

# Tidy the Data

Use the pipe **%>%** symbol **(Ctrl + Shft + M)** to simplify coding. 

### gather() function template

new object <- data %>% 
  gather(key = , value = , columns to gather, other arguments)
  
```{r, the gather function, eval = FALSE}

Ibis2 <- Ibis %>% 
  gather(key = Year, value = Count, `1972`:`1975`, convert = TRUE)

# What does the convert = TRUE argument do?

# Knit the table in a nice format. Notice how you can nest functions with the pipe.

Ibis2 %>% 
  head(n=10) %>% 
  kable(caption="Table 2: Ibis Data in a Long Format") %>%
  kable_styling(full_width = T, bootstrap_options = c("hover", "condensed"))

```

## Task 2:

There are two other ways to identify the the fields to gather in this example. Can you figure this out?  Tip: one way is to identify what not to gather and the other identifies the columns by location.

```{r, Task 2} 


```

## Task 3:

The `spread()` function restructures the long format to a wide format.  Use this function to return our gathered dataset back into the orignal format.  This one is easier than the gather function.

###  spread() template

newobject <- data %>% 
  spread(key = column to pivot, value = column with the values)

```{r, Task 3}  


```

# Export the Data

The example below shows you how to easily export the data as a .txt file. You can also export the data as a .csv file type. 

```{r, write the tables out as a csv, message = FALSE, eval=FALSE}

write.table(Ibis2, file = "IbisClean.txt", row.names = FALSE, sep = "\t")

# row names are output by default, and we don't want that.
# sep = "\t" indicates that the file is to be tab seperated.

```

## Task 4:

Export the data as a .csv file.  Use the Help on the *write.csv()* function to learn how. Make sure you omit row names and include a seperator.

```{r, Task 4}



```


# Independent Exercise

1.  Use the `read_csv()` function to import the *Sage Grouse Data.csv* dataset.  Assign the data into a new object called *Sage*

2.  Examine the data.  How many Observations and variables do we have? What are the variable names?  What data type are the 4 date fields and what should they be?

3. We have multiple observational units combined in one table. How many do you think we have?

4.  Some variables are in wide format (i.e., observations), can you pick them out? 

```{r, Tidy Exercise}

```

# Class Exercise

Question: What is the maximum count of male and female sage grouse per lek?

To answer this question, we'll need to use the `select()`, `group_by()` and `summarise()`functions from the **dplyr** package.  This is a great introduction of what we will be learning more of tomorrow.

## Steps:

1. Select only the variables you need using the `select()` function.  
2. Gather the *FieldResults* fields into a long format using the `gather()` function.  Use the `na.rm = TRUE` argument to remove the cells with missing values.
3. Seperate the field with the data using the `seperate()` function
4. Use the `group_by()` and `summarise()`functions to calculate the maximum number of sage grouse per lek.  Note:  `summarise()` is finiky to missing values, that's why we need to filter them out first. Alternatively, you could just use add `na.rm = TRUE` at the end of the `summarise( )` as well.


```{r, class exercise}


```









